Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-16 Manuscript/manuscript.Rmd
manuscript edits

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases. <!--Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 4).-->

@@

When models were built on randomly oversampled (ROS) data, we found a substantial increase in MCC for the GM12878 cell line (Figure 5 A-C). The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Although performances were low at lower resolutions, MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 10). For K562, a much sharper negative trend was featured for MCCs as resolution decreased, than in the GM12878 cell line.

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, for all 3 predictor types and across resolutions (Figure 5). SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. Here, distance type predictors were found to again yield the highest MCC values across resolutions. SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 10). Moreover, for these sets of data characteristics, we again found that there was a degredation in model performance as data resolution decreased. This was much more pronounced in the K562 cell line. On the GM12878 cell line, there were only marginal differences in performance between 10 kb and 25 kb, followed by a steep decline at 50 kb. Then, the MCCs showed a momentary increase in performance at 100 kb, the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, starting with highest performance at 10 kb and ending with the worst performance at 100 kb. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on distance-type predictors and incorporates both random under- and over-sampling. Furthermore, models built on TAD boundaries called at higher resolutions exhibited better performances.

>>>>

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 5B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 5C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).

@@

When models were built on randomly oversampled (ROS) data for GM12878, we found an increase in MCC compared to no resampling, at higher resolutions. The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Again, distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 3).

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, especially between models with OC and OP predictor types. When using RUS or SMOTE, implemented on distance type predictors, we found a clear improvement compared to the other predictor types. However, there was less of a separation between ROS and RUS techniques within the ensemble of models using distance type predictors.

Overall, SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. Here, distance type predictors were found to again yield the highest MCC values across resolutions. It should be noted that, SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 3). Moreover, for these sets of data characteristics, we again found that there was a degredation in model performance as data resolution decreased. Results indicated that performances exhibited much more variability when built on lower resolutions, specifically at the 100 kb level. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.
